---
title: "R exercise: phenotype data proprocessing"
author: "Haoyue Shuai"
date: "Oct. 8th, 2020"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '4'
  pdf_document:
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
#please do not touch this chunk
knitr::opts_chunk$set(echo = TRUE, results = "hold",fig.width = 7, fig.height = 4)
if(!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, ggplot2, plyr, tidyverse, pander, ggpubr, rapportools, knitr, pROC, reshape2)  
```

\newpage

## Instruction

This exercise was adapted by Haoyue Shuai from one of her analysis on phenotype data in preparation for genetic association studies. Don't panic if you're just starting to learn R. The exercise does not involve writing serious software programs in R, but rather to execute some interactive and intuitive commands
to get some preliminary phenotype data analysis done. There will be some questions throughout that you can answer using your own words (and some data science intuition) with a few sentences. Some of them involve writing additional codes but you can mostly find and modify codes we provide as examples to address the questions. You do not have to answer all the questions but we strongly encourage you to make an attempt.

Please save your work as you complete them, but **do not** push your version of the file back to your forked github repo. Instead please follow the instructions in Task 4 to report the analysis back to us.

### Dataset 

The data-set can be found as `data/UKB_Phenotype/data_cleaned.csv`.

\newpage

\newpage 

## Genetic association study

We perform genetic association studies to identify genetic factors (variants) that may be involved in a complex trait (tinnitus, asthma, etc) etiology.
In brief, genetic association studies compare and identify difference in genetic data of individuals with disease (cases) to those without (controls). 
We report genetic variants that are observed more frequently in cases than in controls.

In order to perform genetic association studies, we need phenotype data and genotype data from individuals we collect.

- Phenotype and covariate data: age, sex, height, weight, condition for that trait (tinnitus in the example below case), etc.
- Genotype data: You can roughly understand it as a sequence of the bases in DNA molecules, A/T/C/G, for all chromosomes in human genome.

### Disease phenotype data 

We use a toy data-set extracted from the UK Biobank project.

**Load the data**

Load `UKB_Phenotype/data_cleaned.csv`. Note, you only need to complete the codes when seeing `YOUR CODE`. Please execute other existing codes as is.

```{r}
# you need to put the data-set in the same folder
# where this .rmd file sits,
# which is here:
getwd()
sub_UKBB<-read.csv("data_cleaned.csv")
```

**Exploratory data analysis (EDA) of the data**
```{r}
dim(sub_UKBB) # This data covering 144756 participants and 11 variables of them (IID, FID, etc)
```

```{r}
colnames(sub_UKBB) # 11 variables 
```

```{r}
summary(sub_UKBB) 
```

```{r}
head(sub_UKBB) # show part of the data
```

**Q1:** How many females and males are there in this data? Please show your code below how these numbers are computed.

**A1.** There are 77535 females and 67221 males in this data. (please see code below)

```{r}
# Number per gender
table(sub_UKBB$f.22001.0.0)
```


**Q2:** What type of trait do you think best describes tinnitus as a phenotype? 
A.Binary B. Continuous C. Ordinal 

**A2:** Tinnitus is a qualitative trait, and can be either present (yes) or absent (no) in an individual. Since it can only take two values, it can be best described as a binary trait.


**Q3:** Recode f.4803

Field 4803 (f.4803) is the answers from participants for ACE touchscreen question "Do you get or have you had noises (such as ringing or buzzing) in your head or in one or both ears that lasts for more than five minutes at a time?" 

These fields contains answers to the questions in their first, 2nd, 3rd and 4th hospital visit: f.4803.0.0, f.4803.1.0, f.4803.2.0, f.4803.3.0. 

```{r recode}
# Recode function:
recode<-function(df,column_name){
  new_names<-c()
  for (i in column_name){
    new_column_name<-paste0(i,"_recode")
    new_names<-c(new_names,new_column_name)
    df[,new_column_name] <- revalue(df[,i], c("No, never"= 0, 
                                            "Yes, but not now, but have in the past"= 1, 
                                            "Yes, now some of the time"= 1, 
                                            "Yes, now a lot of the time"= 1,
                                            "Yes, now most or all of the time"= 1,
                                            "Prefer not to answer"= NA,
                                            "Do not know"= NA ))
  }
  return (list(df=df,new_column_names=new_names))
}

# columns needs to be recoded:
column_name<-c("f.4803.0.0","f.4803.1.0","f.4803.2.0","f.4803.3.0")

# get a new data.frame with recoded columns added:
df_recode<-recode(df=sub_UKBB,column_name)$df

# get names of recoded columns:
new_column_names<-recode(df=sub_UKBB,column_name)$new_column_names

# show recode summary:
for (i in new_column_names)
{cat(i,"summary:");print(table(df_recode[,i]));cat("\n")}
```

What do you think has been achieved by recoding these fields?

**A3:** By recoding these fields, we have computed for each hospital visit the number of individuals who reported to have tinnitus (1), and the number of individuals who did not (0).


**Q4:** Define case and control status of tinnitus for each participant in the study:

```{r}
data_sub <- df_recode[,new_column_names]
# Function to define cases
f<-function(x){
  visit<-c()
  for (i in 1:4){
    if (!is.na(x[i]))
    {visit<-c(visit,x[i])}
  }
  if ("1" %in% visit){result= TRUE}
  else{result=FALSE}
  return (result)
}
# Apply the above function
df_recode$cases<-apply(data_sub, 1, f)
head(df_recode,10)
```

How many cases and how many controls do we have for this phenotpype?

**A4:** There are 43206 cases (TRUE) and 101550 (FALSE) controls for this phenotype. (please see code below)

```{r}
# Number of cases and controls
table(df_recode$cases)
```


**Q5:** Extract a subset of columns from all participants for association study. 

```{r}
df_cases  <- df_recode %>%
  select(IID,FID,cases)%>% 
  filter(cases==TRUE)
head(df_cases,10)  
```
Please modify codes above to extract all the controls and keep only these columns: `FID`, `IID`, `cases`, `f.22001.0.0`, `f.21003.0.0`, `f.21003.1.0`, `f.21003.2.0`, `f.21003.3.0`. Please show the first 10 rows of the output.
```{r}
# YOUR CODE
df_controls  <- df_recode %>%
  select(IID,FID,cases,f.22001.0.0,f.21003.0.0,f.21003.1.0,f.21003.2.0,f.21003.3.0)%>% 
  filter(cases==FALSE)
head(df_controls,10)
```
\newpage

### Covariates
 
**Q6:** Field 21003 contains the information of the age of participants, same as field 4803. Note that some of them have more than one age. Can you guess why?

**A6:** Some participants have more than one age because they had different age at each hospital visit.


**Q7:** For those with more than one age records, which age do you think should be used in the genetic association analysis?

**A7:** We should use the most recent age that corresponds best to the participant's status (case vs control). e.g. for a participant who reported to have tinnitus (1) at one of his hospital visits, the age to use is the one recorded at that specific hospital visit.


**Q8:** Please compute a summary of age information for controls (you can use `summary()` function in R):
```{r}
#YOUR CODE
df_agecontrols <- df_controls %>%
  select(f.21003.0.0,f.21003.1.0,f.21003.2.0,f.21003.3.0)
summary(df_agecontrols)
```
\newpage

### Association testing via regression analysis

To identify genetic factors that may be involved in this trait (tinnitus), we would need to find the association between the genotype and the phenotype. 
Regression analysis is the basis of many association analysis. Instead of overwhelming you with huge genotype data, we use here a simple dataset for regression analysis to demonstrate what association studies look like.


We fit below simple linear model with 2 variables from a data-set to see their relationship. For example `mpg` vs. `weight` in this `ISLR::Auto` data-set. 

**Q9:** Is there association between `mpg of the car` and `weight of the car`? If so, it appearing to be positive or negative? Is the association significant and why?

**A9:** Since the beta coefficient (or the slope; -0.007647) is different from zero and has a negative sign, we can say that there is a negative association between the mpg and the weight of the car. The p-value of the beta coefficient is nearly equal to zero (<2e-16), suggesting a significant association between mpg and weight.

```{r, echo=FALSE}
# check if you have ISLR package, if not, install it
if(!requireNamespace('ISLR')) install.packages('ISLR') 
auto_data <- ISLR::Auto
#  fit a linear regression model
fit_1<-lm(mpg ~ weight, auto_data)
summary(fit_1)
```

**Q10:** Please create a new variable to indicate cars having MPG greater than 23 as 1, otherwise 0, then use logistic regression via `glm()` function to analyze association between weight and this new variable you just created. Please comment on what you find.

**A10:** Using the logistic regression model, the estimate of the beta coefficient (-0.0040954) is different from zero, and the corresponding p-value (<2e-16) is less than <0.05. This indicates that there is a significant association between the weight the mpg of the car. (please see code below)

**Q11:** Find the `Estimates` from your association results summary. How do you interpret the estimated effects of weight in the context of linear regression, and in the context of logistic regression?

**A11:** In the context of linear regression, we can say that for every 1 unit increase in the weight of the car, the mpg is reduced by up to 0.007647 unit. While in the context of logistic regression, 1 unit increase on the weight of the car will decrease the odd of mpg being more than 23 by exp(0.0040954) 0.99 time.

```{r}
#YOUR CODE
# Extract colums to be used
newauto_data <- auto_data %>%
  select(mpg,weight)
# Recode mpg
newauto_data$mpg <- ifelse(test = newauto_data$mpg >23, yes = 1, no = 0)
# fit a logistic regression model
fit_2 <- glm(mpg ~ weight, data = newauto_data, family = binomial)
summary(fit_2)
```
